Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2023-02-10 analysis/paper/paper.qmd
editing up to the discussion

@lee2019were first applied TCSA to stemmed points from Korea along with use-wear analysis. They analyzed stemmed points from Jingeuneul, located in southwest of Korea, which has the largest number of stemmed points (n = 99) found so far in a single site. For TCSA, they were only able to use 10 stemmed points because they needed to select the stemmed points that retained the widest and thickest part of the specimens for TCSA metric and also displayed diagnostic impact fractures for the use-wear analysis. The purpose of using TCSA in their research was to compare the values to North American dart tips and arrowheads. Their results show that the TCSA range of Jingeuneul stemmed points is relatively wide, overlapping with both North American dart tips and arrowheads. According to their use-wear analysis, a significant number of the stemmed points have diagnostic impact fractures (DIFs) on the surface, likely caused by longitudinal forces from the shaft. They conclude that stemmed points may have been used as spear-throwers or bows. Inspired by @lee2019were and TCSA research from lithic assemblages in other parts of the world, we aim to investigate TCSA values for stemmed points from all over South Korea to better understand their likely uses during the Late Paleolithic.

@@

After the first discovery at the Suyanggae site, more than 450 stemmed points have been found in over 30 sites throughout South Korea [@Sohn_1967; @lee2019were; @Chong_2021]. Most sites contain only a few stemmed points and only a few sites have many more, such as Jingeuneul, Suyanggae (n = 55), and Yongsandong (n = 38) [@Kim_2017]. Among these stemmed points, we selected those that retain their widest and thickest parts. We selected stemmed points that were found during field surveys as well as those were from sites that were never dated but were associated with other Late Palaeolithic artifacts. Applying these sampling criteria resulted in a sample of `r total_sp` stemmed points from `r total_ass` assemblages unearthed from `r total_site` sites spanning the period 44-10 ka (@fig-map). The dimensions of the `r total_sp` stemmed points were obtained from published excavation reports and by direct measurements during our visits to the collections of local museums and archaeological institutions in Korea.

@@

To aid with interpreting our results, we followed the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths and calibrating existing ranges (@tbl-tcsa-ranges). We excluded `r n_omit` artifacts from our dataset with TCSA greater than 250 and thus outside of the range of the comparative data. For this study we include TCSA values from a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites.

@@

Different sizes of stone artifacts can constrain or enable different functions. Overall size of stone points has been used as a potential proxy to distinguish different armature types [@thomas1978arrowheads; @sahle2019assessment]. We chose weight as a proxy for overall size of the stemmed points. Weight is a reliable discriminator between tools of different sizes and a descriptive attribute which can be measured rapidly and objectively and serve as a criterion of likeness or dissimilarity in projectile point description [@fenenga1953weights; @shea2006origins]. We then explored the relationship between weight and raw materials on TCSA values. Excluding points that we were not able to measure or obtain the records of their weight, we explored the relationship between TCSA and weight for `r n_size` artifacts.

@@

We summarized the distribution of TCSA values for each assemblage to examine spatial patterns in TCSA values. Among `r n_sites` artifacts, we combined sites that contains fewer than 5 stemmed points aside and named them as "Other". We explored possible effect of environmental variation on TCSA values by grouping stemmed points by the eco-regional zones that they were found in, and comparing the distributions of TCSA values across the different zones. We compared distributions across four vegetation zones, and also across 16 zones of different geographical characteristics.

@@

We predict that TCSA values vary across vegetation types and geographical zones in response to heterogeneously distributed food resources. In harsh environments with lower patch productivity, similar to the case of LGM duration, we predict a further increase in the variability of TCSA values. On the other had, we predict a reduction in the range of TCSA values in affluent patches with predictable types of prey. In their study of hunter-gatherer mobility strategies, @hamilton2016ecological show that resources are abundant and predictable along coasts or lake shores, which is highly dependent on temperature and precipitation and thus hunter-gatherers often become effectively sedentary. Our prediction is that the range of TCSA values will be narrower in coastal areas such as ST and SW of vegetation zones and ESC of geographical zones than in inland areas such as CT of vegetation zones and CI of geographical zones.

The entire R code [@core2021team] and data files used for all the analysis and visualizations contained in this paper is openly available at https://doi.org/10.17605/osf.io/dqna8 to enable re-use of materials and improve reproducibility and transparency [@marwick2017computational]. All of the figures, tables, statistical test results presented here in addition to the raw data can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.

@@

Using weight as a size proxy we examined the relationship between size and likely use of the stemmed points inferred from TCSA values. We conducted a univariate cluster analysis [@song2020efficient; @wang2011ckmeans] of stemmed points by weight and it revealed three clusters of artifact sizes (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'Weight(g)'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'Weight(g)'), digits=1)`). Cluster 1, the smallest (lightest) artifacts, have a lower range of TCSA values, compared to Cluster 2 (B of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'min_tcsa'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'min_tcsa'), digits=1)`). TCSA values for Cluster 3 are the highest, except for one artifact, which is lower than 50. There are a few overlaps, but we can still assume that artifacts in different size clusters might be made for different purposes.

@@

@fig-tcsa-raw-materials shows that about half of the stemmed points were made from shale and its TCSA range is widely distributed. Acidic volcanic rocks tend to have skewed lower TCSA values. The other raw materials including hornfels, rhyolite, and tuff show a wide distribution, similar to shale. The category of "Other" raw materials include porphyry, trachyte, felsite, chert, quartz, quartzite, granite, mudstone, and unidentified rocks. Overall, there is no clear pattern of TCSA among raw materials.

@@

@fig-tcsa-size shows a positive relationship between TCSA, raw material and artifact weight. There is a stronger correlation between TCSA and artifact weight for acidic volcanic points (i.e. points are closer to the regression line), whereas the correlation for shale stemmed points is weaker. The other raw materials show various distribution patterns around the regression line, confirming that raw material appears not to have strongly influenced TCSA (see also @fig-tcsa-raw-materials).

@@

We computed artifact TCSA values for `r n_sites` sites that contain more than 5 stemmed points to observe variation between sites (@fig-tcsa-per-sites). Sites with fewer than 5 stemmed points were grouped under the category of "Other". Among the sites in our sample, Suyanggae has the most stemmed points and the widest range of TCSA values. This suggests that people made stemmed points for accomplishing a variety of tasks at Suyanggae site. Nosanri, Sibuk, and Wolpyeng have a narrower range of lower values, which indicates less diverse likely uses for stemmed points. Yongsujaeul shows a narrower range but higher TCSA value. Yongsandong has the second most stemmed points and shows roughly two different clusters of TCSA values.

@@

In order to explore spatial patterns and the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional zones. Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of vegetation map or Central inland (CI) (@fig-gis-tcsa-sites panel A & D) of eco-region map. We observed that there are statistically significant differences in the distributions of TCSA values between some zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`). The Tukey's HSD results show that TCSA values from the South Temperature Zone (ST) zone are significantly different from the other two vegetation zones (@fig-gis-tcsa-sites panel C). Looking into this further, we see that there are only six artifacts from the ST zone, with only three of these having TSCA values above 175. The other three have TCSA values lower than 100, similar to the other zones (@fig-gis-tcsa-sites panel B). Given the small number of artifacts with extreme TCSA values in the ST zone, we hesitate to conclude this is an archaeologically significant pattern of different TCSA values. The Tukey HSD pairwise differences show no pairs of eco-regions with significant differences in TCSA values (@fig-gis-tcsa-sites panel F). Overall, our results indicate no clear patterns in the distribution of TCSA values across the vegetation and eco-regional zones.

>>>>

@lee2019were first applied TCSA to stemmed points from Korea along with use-wear analysis. They analyzed stemmed points from Jingeuneul, located in southwest of Korea, which has the largest number of stemmed points (n = 99) found so far in a single site. For TCSA, they were only able to use 10 stemmed points because they chose to select the stemmed points that retained the widest and thickest part of the specimens for TCSA metric and also displayed diagnostic impact fractures for the use-wear analysis. The purpose of using TCSA in their research was to compare the values to North American dart tips and arrowheads. Their results show that the range of TCSA values for Jingeuneul stemmed points is relatively wide, overlapping with both North American dart tips and arrowheads. According to their use-wear analysis, a significant number of the stemmed points have diagnostic impact fractures (DIFs) on the surface, likely caused by longitudinal forces from the shaft. They conclude that stemmed points may have been used as spear-throwers or bows. Inspired by @lee2019were and TCSA research from lithic assemblages in other parts of the world, we aim to investigate TCSA values for a much larger sample of stemmed points from all over South Korea to better understand their likely uses during the Late Paleolithic.

@@

After the first discovery at the Suyanggae site, more than 450 stemmed points have been found in over 30 sites throughout South Korea [@Sohn_1967; @lee2019were; @Chong_2021]. Most sites contain only a few stemmed points and only a few sites have many more, such as Jingeuneul, Suyanggae (n = 55), and Yongsandong (n = 38) [@Kim_2017]. Among these stemmed points, we selected those that retained their widest and thickest parts. We selected stemmed points that were found during field surveys as well as those were from sites that were never dated but were associated with other Late Palaeolithic artifacts. Applying these sampling criteria resulted in a sample of `r total_sp` stemmed points from `r total_ass` assemblages unearthed from `r total_site` sites spanning the period 44-10 ka (@fig-map). The dimensions of the `r total_sp` stemmed points were obtained from published excavation reports and by direct measurements during our visits to the collections of local museums and archaeological institutions in Korea.

@@

To aid with interpreting our results, we followed the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing the analysis by @wadley2008segment of backed microliths (@tbl-tcsa-ranges). We excluded `r n_omit` artifacts from our dataset with TCSA greater than 250 and thus outside of the range of the comparative data. For this comparison with weapon-delivery systems we include TCSA values from a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites.

@@

Different sizes of stone artifacts can constrain or enable different functions. Overall size of stone points has been used as a potential proxy to distinguish different armature types [@thomas1978arrowheads; @sahle2019assessment]. We chose weight as a proxy for overall size of the stemmed points. Weight is a reliable discriminator between tools of different sizes and a descriptive attribute which can be measured rapidly and objectively [@fenenga1953weights; @shea2006origins]. We then explored the relationship between weight and raw materials on TCSA values. Excluding points that we were not able to directly measure or obtain the records of their weight, we explored the relationship between TCSA and weight for `r n_size` artifacts.

@@

We summarized the distribution of TCSA values for each assemblage to examine spatial patterns in TCSA values. Among `r n_sites` sites, we combined those that contain fewer than five stemmed points and named them as "Other". We explored possible effect of environmental variation on TCSA values by grouping stemmed points by the eco-regional zones that they were found in, and comparing the distributions of TCSA values across the different zones. We compared distributions across four vegetation zones, and also across 16 zones of different geographical characteristics.

@@

We predict that TCSA values vary across vegetation types and geographical zones in response to heterogeneously distributed food resources. In harsh environments with lower patch productivity, similar to the case of LGM duration, we predict a further increase in the variability of TCSA values. On the other had, we predict a reduction in the range of TCSA values in affluent patches with predictable types of prey. In their study of hunter-gatherer mobility strategies, @hamilton2016ecological show that resources are abundant and predictable along coasts or lake shores, which is highly dependent on temperature and precipitation and thus hunter-gatherers often become effectively sedentary. Our prediction is that the range of TCSA values will be narrower in coastal areas such as vegetation zones ST and SW and geographical zone ESC, compared to inland areas such as vegetation zone CT and geographical zone CI.

The entire R code [@core2021team] and data files used for all the analysis and visualizations contained in this paper is openly available at https://doi.org/10.17605/osf.io/dqna8 to enable re-use of materials and improve reproducibility and transparency [@marwick2017computational]. All of the figures, tables, statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.

@@

Using weight as a size proxy we examined the relationship between size and likely use of the stemmed points inferred from TCSA values. We conducted a univariate cluster analysis [@song2020efficient; @wang2011ckmeans] of stemmed points by weight, revealing three clusters of artifact sizes (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'Weight(g)'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'Weight(g)'), digits=1)`). Cluster 1, the smallest (lightest) artifacts, has a lower range of TCSA values, compared to Cluster 2 (B of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'min_tcsa'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'min_tcsa'), digits=1)`). TCSA values for Cluster 3 are the highest, except for one artifact, which is lower than 50. There are a few overlaps, but we can still assume that artifacts in different size clusters might have been made for different purposes.

@@

@fig-tcsa-raw-materials shows that about half of the stemmed points were made from shale and its TCSA range is widely distributed. Acidic volcanic rocks tend to have skewed lower TCSA values. The other raw materials including hornfels, rhyolite, and tuff show a wide distribution, similar to shale. The category of "Other" raw materials include porphyry, trachyte, felsite, chert, quartz, quartzite, granite, mudstone, and unidentified rocks. Overall, there is no clear pattern of TCSA values among different raw materials.

@@

@fig-tcsa-size shows positive relationships between TCSA, raw material and artifact weight. There is a stronger correlation between TCSA and artifact weight for acidic volcanic points (i.e. points are closer to the regression line), whereas the correlation for shale stemmed points is weaker. The other raw materials show various distribution patterns around the regression line, confirming that raw material appears not to have strongly influenced TCSA (see also @fig-tcsa-raw-materials).

@@

We computed artifact TCSA values for `r n_sites` sites that contain more than five stemmed points to observe variation between sites (@fig-tcsa-per-sites). Sites with fewer than five stemmed points were grouped under the category of "Other". Among the sites in our sample, Suyanggae has the most stemmed points and the widest range of TCSA values. This suggests that people made stemmed points for accomplishing a variety of tasks at Suyanggae site. Nosanri, Sibuk, and Wolpyeng have a narrower range of lower values, which indicates less diverse likely uses for stemmed points. Yongsujaeul shows a narrower range but higher TCSA values. Yongsandong has the second highest number of stemmed points and shows two clusters of TCSA values.

@@

In order to explore spatial patterns and the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional zones. Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of the vegetation map or Central inland (CI) (@fig-gis-tcsa-sites panel A & D) of the eco-region map. We observed that there are statistically significant differences in the distributions of TCSA values between some zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`). The Tukey's HSD results show that TCSA values from the South Temperature Zone (ST) zone are significantly different from the other two vegetation zones (@fig-gis-tcsa-sites panel C). Looking into this further, we see that there are only six artifacts from the ST zone, with only three of these having TSCA values above 175. The other three have TCSA values lower than 100, similar to the other zones (@fig-gis-tcsa-sites panel B). Given the small number of artifacts with extreme TCSA values in the ST zone, we hesitate to conclude this is an archaeologically significant pattern of different TCSA values. The Tukey HSD pairwise differences show no pairs of eco-regions with significant differences in TCSA values, likely due to the small number of points in several zones (@fig-gis-tcsa-sites panel F). Overall, our results indicate no clear patterns in the distribution of TCSA values across the vegetation and eco-regional zones.
